---
title: 'Health and Economic Outcomes of Weather Events 1951-2011: An Overview'
author: "ISY"
date: "12/21/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Synopsis

Storms and other severe weather events can cause both public health and economic 
problems for communities and municipalities. In this research, data from US National
Oceanic and Atmospheric Administration had been obtained and processed to see the
outcomes by each event. Data were obtained (link in loading and processing the raw
data section), cleaned (data processing section) and summarized (pre-analysis section).
Results have been shown in a respective section. Data were analysed in RStudio
R Core Team (2020). R: A language and environment for statistical computing. R
  Foundation for Statistical Computing, Vienna, Austria. URL
  https://www.R-project.org/.



## Loading and Processing the Raw Data

From the [EPA Air Quality System](http://www.epa.gov/ttn/airs/airsaqs/detaildata/downloadaqsdata.htm), we
obtained data particulate matter air pollution levels that is monitored accros the
USA by a nationwide PM monitoring source. We obtained files that presents the data
from 1999 and 2012.

### Data Processing

Dataset were given in [this](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) link in the assignment instructions. 

```{r}
#Set directories and download dataset

directory <- getwd()
urldata <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
filepath <- paste0(directory,"/stormdata.csv.bz2")

if (!file.exists(filepath)) {download.file(urldata, destfile = filepath)}

#Download Data Documentation
urldoc <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
docpath <- paste0(directory,"/datadoc.pdf")
if (!file.exists(docpath)) {download.file(urldoc, destfile = docpath)}

storm <- read.csv(file = filepath, na.strings = NA)

#Summarize the dataset

str(storm)
```

### Pre-analizing Processing

These are the questions that are asked in the assignment:

1. Across the United States, which types of events (as indicated in the 
**EVTYPE** variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic 
consequences?

Since our dataset has a relatively big size, required information will be selected
from the dataset and a new dataset will be created. The information should be on
health and economic assets.

As we check from the data documentation, the following columns in *storm* dataset
contains information to regarding:

*EVTYPE: Event Type
*FATALITIES : Mortality
*INJURIES : # of Injuries
*PRODMG : Property Damage in US Dollars
*PROPDMGEXP: The units property damage
*CROPDMG : Crop Damage in US Dollars
*CROPDMGEXP: The units for property damage

*dplyr*, *data.table* and  *lubridate* packages will be used for this pre-analizing
process. 

```{r}
library(dplyr)
library(data.table)
library(lubridate)

vars <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP","CROPDMG","CROPDMGEXP")
mydata <- storm[, vars]

```

Check for information on missing values.

```{r}
for (i in colnames(mydata)) {print(sum(is.na(mydata[[i]])))}
```

Event types might vary, We wil check for any combination possibilities.

```{r}
sort(table(mydata$EVTYPE), decreasing = TRUE)[1:20]
```

We will rename the event types to their main categories in terms of presentation
purposes. If events are relatively small in numbers, we will assign them as "Other".
Regrouping will be done by extracting values to a new variable: "EVENT". 
Note that we will focus on storm related info.

```{r}

mydata$EVENT <- "Other"

mydata$EVENT[grep("HAIL", mydata$EVTYPE, ignore.case = TRUE)] <- "Hail"
mydata$EVENT[grep("HEAT", mydata$EVTYPE, ignore.case= TRUE)] <- "Heat"
mydata$EVENT[grep("FLOOD", mydata$EVTYPE, ignore.case = TRUE)] <- "Flood"
mydata$EVENT[grep("WIND", mydata$EVTYPE, ignore.case = TRUE)] <- "Wind"
mydata$EVENT[grep("STORM", mydata$EVTYPE, ignore.case= TRUE)] <- "Storm"
mydata$EVENT[grep("SNOW", mydata$EVTYPE, ignore.case = TRUE)] <- "Snow"
mydata$EVENT[grep("TORNADO", mydata$EVTYPE, ignore.case = TRUE)] <- "Tornado"
mydata$EVENT[grep("WINTER",mydata$EVTYPE, ignore.case= TRUE)] <- "Winter"
mydata$EVENT[grep("RAIN", mydata$EVTYPE, ignore.case = TRUE)] <- "Rain"

#Check for newly created variable.

sort(table(mydata$EVENT), decreasing = TRUE)

```

Check for distinct values for expedition data. 

```{r}
for (i in c("PROPDMGEXP","CROPDMGEXP")) {print(unique(mydata[[i]]))}
```

Some of these values mean:

*K: Thousand dollars (10^3)
*B: Billion dollars (10^9)
*M: Million dollars (10^6)

Rest will be considered as dollars.

```{r}
#Check for damage variables class. If anything different than character, convert to character

mydata$PROPDMGEXP <- as.character(mydata$PROPDMGEXP)
mydata$CROPDMGEXP <- as.character(mydata$CROPDMGEXP)

#Change property expedition values to numeric, multiply the new values with 
#property damage value and store the new values in a new variable.

mydata$PROPDMGEXP[grep("K", mydata$PROPDMGEXP, ignore.case = TRUE)] <- 3
mydata$PROPDMGEXP[grep("M", mydata$PROPDMGEXP, ignore.case = TRUE)] <- 6
mydata$PROPDMGEXP[grep("B", mydata$PROPDMGEXP, ignore.case = TRUE)] <- 9
mydata$PROPDMGEXP[grep("K|M|B", mydata$PROPDMGEXP, ignore.case= TRUE)] <- 0

mydata$PROPDMGEXP <- as.numeric(mydata$PROPDMGEXP)
mydata$PROPDMGEXP[is.na(mydata$PROPDMGEXP)] <- 0
mydata$PROP.DMG <- mydata$PROPDMG * 10^mydata$PROPDMGEXP

```

```{r}
# Change crop damage expeditions to numeric, multiply the new values with crop damage
#values and store the new values in a new variable

mydata$CROPDMGEXP[grep("K", mydata$CROPDMGEXP, ignore.case = TRUE)] <- 3
mydata$CROPDMGEXP[grep("M", mydata$CROPDMGEXP, ignore.case = TRUE)] <- 6
mydata$CROPDMGEXP[grep("B", mydata$CROPDMGEXP, ignore.case = TRUE)] <- 9
mydata$CROPDMGEXP[grep("K|M|B", mydata$CROPDMGEXP, ignore.case = TRUE)] <- 0

mydata$CROPDMGEXP <- as.numeric(mydata$CROPDMGEXP)
mydata$CROPDMGEXP[is.na(mydata$CROPDMGEXP)] <- 0
mydata$CROP.DMG <- mydata$CROPDMGEXP * 10^mydata$CROPDMGEXP

#Check the new variables

str(mydata$PROP.DMG)
str(mydata$CROP.DMG)
```


Check for any NA's in the dataset

```{r}
for (i in colnames(mydata)) {print(sum(is.na(mydata[[i]])))}
```

Since we need the the outcomes for both health and economics, we will create a
data frame that contains the summary of both health and economics
```{r}
outcomes <- mydata %>% group_by(EVENT) %>% 
     summarize(Fatalities = sum(FATALITIES),
               Injuries = sum(INJURIES),
               totalHealth = sum(sum(FATALITIES), sum(INJURIES)),
               PropDMG = sum(PROP.DMG),
               CropDMG= sum(CROP.DMG),
               totalEcon = sum(sum(PROP.DMG),sum(CROP.DMG)))

# Add total cases
totalCases <- as.data.frame(table(mydata$EVENT))[,2]
outcomes <- cbind(outcomes, totalCases)

#Order the columns for better exploration
outcomes <- outcomes[, c(1,8,2:7)]

```


## Analysis

### Part 1: Public Health Outcomes

Question is: Across the United States, which types of events (as indicated in the 
**EVTYPE** variable) are most harmful with respect to population health?

We have created an *outcomes* table before. We will create a barplot to show the
health outcomes. *ggplot2* package will be used.

```{r}
library(ggplot2)
plot1  <- ggplot(data = outcomes, aes(x=totalHealth, y=EVENT), ) + 
     geom_bar(stat="identity") +
     xlab("Total Health Outcome") +
     ylab("Type of Weather Event") +
     ggtitle("Total Health Outcomes of Weather Events 1951-2011") +
     theme(plot.title = element_text(hjust = 0.5))

plot1

```

As a result, worst health outcomes had been seen because of tornados.  

### Part 2: Economic outcomes

Question is: Across the United States, which types of events have the greatest 
economic consequences?

*outcomes* dataset will be used.

```{r}
plot2 <- ggplot(data= outcomes, aes(x= totalEcon/1E9, y= EVENT)) + 
     geom_histogram(stat = "identity") + 
     xlab("Economic Loss (in billion USD)") +
     ylab("Type of Weather Event") +
     ggtitle("Total Economic Outcomes of Weather Events 1951-2011") +
     theme(plot.title = element_text(hjust = 0.5)) 

plot2
```

As a result, we can say that most devastating event for property and corps is
flood. (excluding "others")

